quollr: An R Package for Visualizing 2-D Models from Non-linear Dimension Reductions in High Dimensional Space

### 2025-11-08 paper-quollr.tex
Changed the notations to view in HTML as well

This paper presents the R package, \texttt{quollr}, which is useful for understanding how NLDR warps high-dimensional space and fits the data. Starting with an NLDR layout, our approach is to create a $2\text{-}D$ wireframe model representation, that can be lifted and displayed in the high-dimensional ($p\text{-}D${}) space (Figure \ref{fig:overview}).

\begin{figure}[H]
\includegraphics[width=1\linewidth]{paper-quollr_files/figure-latex/overview-1} \caption{Wireframe model representation of the NLDR layout, lifted and displayed in high-dimensional space. The left panel shows the NLDR layout with a triangular mesh overlay, forming the wireframe structure. This mesh can be lifted into higher dimensions and projected to examine how the geometric structure of the data is preserved. Panels (a1–a4) display different $2\text{-}D$projections of the lifted wireframe, where the underlying curved sheet structure of the data is more clearly visible. The triangulated mesh highlights how local neighborhoods in the layout correspond to relationships in the high-dimensional space, enabling diagnostics of distortion and preservation across dimensions.}\label{fig:overview}
\end{figure}

@@

The package is organized into seven core modules corresponding to stages of the analysis workflow: preprocessing, $2\text{-}D$ model construction, lifting into $p\text{-}D$, prediction, error computation, visualization, and interactivity. Each module performs a distinct task and communicates through data objects.

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  Data preprocessing: The function \texttt{gen\_scaled\_data()} standardizes the embedding data, manage variable naming, and ensure consistent identifiers across high-dimensional and embedded datasets.
\item
  Construct $2\text{-}D$ model: A series of functions \texttt{hex\_binning()}, \texttt{merge\_hexbin\_centroids()}, \texttt{tri\_bin\_centroids()}, \texttt{gen\_edges()}, and \texttt{update\_trimesh\_index()} generate the hexagonal grid, compute bin centroids, and connect the triangular mesh that defines local neighborhoods in the $2\text{-}D$ space.
\item
  Lift the model into $p\text{-}D$: The function \texttt{avg\_highd\_data()} computes the average of the high-dimensional variables for each bin, linking the $2\text{-}D$ representation back to the original data space.
\item
  Prediction: The function \texttt{predict\_emb()} estimates the embedding of new high-dimensional observations based on the fitted model.
\item
  Error computation: The \texttt{glance()} and \texttt{augment()} function summarizes model performance by comparing the predicted and original embeddings.
\item
  Visualization: Functions such as \texttt{geom\_hexgrid()}, \texttt{geom\_trimesh()}, and \texttt{show\_langevitour()} provide tools for exploring the fitted models through static and dynamic visualizations.
\item
  Interactivity: The functions \texttt{comb\_all\_data\_model()} and \texttt{show\_link\_plots()} generate interactive linked visualizations that connect the $2\text{-}D$ NLDR layout, the corresponding tour view, and the fitted model. Similarly, \texttt{comb\_all\_data\_model\_error()} and \texttt{show\_error\_link\_plots()} integrate the error distribution with the $2\text{-}D$ embedding and dynamic high-dimensional view, enabling interactivity across multiple visual components.
\end{enumerate}

@@

\begin{itemize}
\item
  \texttt{scaled\_nldr\_obj}: the output of \texttt{gen\_scaled\_data()}, which rescales the embedding to the range \([0, 1] \times [0, y_{2,\text{max}}]\), where \(y_{2,\text{max}} = r_2 / r_1\) is the ratio of the embedding ranges. It includes the scaled coordinates (\texttt{scaled\_nldr}) and the original limits (\texttt{lim1}, \texttt{lim2}).
\item
  \texttt{hex\_bin\_obj}: the object created by \texttt{hex\_binning()}, which defines the structure of the two-dimensional hexagonal grid used in modeling. It includes the grid spacing (\texttt{a1}, \texttt{a2}), the number of bins along each axis, the centroids of all hexagons, polygon coordinates for plotting, and the mapping of each data point to its assigned hexagon.
\item
  \texttt{highd\_vis\_model}: the main model object returned by \texttt{fit\_highd\_model()}. It stores all components of the fitted visualization model, including the scaled NLDR data (\texttt{nldr\_scaled\_obj}), the hexagonal bin structure (\texttt{hb\_obj}), the averaged $p\text{-}D$ summaries for each bin (\texttt{model\_highd}), the corresponding $2\text{-}D$ bin centroids (\texttt{model\_2d}), and the triangulated mesh connecting neighboring bins (\texttt{trimesh\_data}).
\end{itemize}

@@

Several core computations within \texttt{quollr} are optimized using compiled C++ code via the \texttt{Rcpp} and \texttt{RcppArmadillo} packages. While the user interacts with high-level R functions, performance-critical steps such as nearest-neighbor searches (\texttt{compute\_highd\_dist()}), error metrics (\texttt{compute\_errors()}), $2\text{-}D$ distance calculations (\texttt{calc\_2d\_dist\_cpp()}), and generation of hexagon coordinates (\texttt{gen\_hex\_coord\_cpp()}) are handled internally in C++. This design provides significant speedups when analyzing large datasets while maintaining a user-friendly R interface. These C++ functions are not exported but are bundled within the package and fully accessible for inspection in the source code.

@@

This function requires several parameters: the high-dimensional data (\texttt{highd\_data}), the emdedding data (\texttt{nldr\_data}), the number of bins along the x-axis (\texttt{b1}), the buffer amount as a proportion of data (\texttt{q}), and benchmark value to extract high density hexagons (\texttt{hd\_thresh}). The function returns an object of class \texttt{highd\_vis\_model} containing the scaled NLDR object (\texttt{nldr\_scaled\_obj}) with three elements: the first is the scaled NLDR data (\texttt{scaled\_nldr}), and the second and third are the limits of the original NLDR data (\texttt{lim1} and \texttt{lim2}); the hexagonal object (\texttt{hb\_obj}), the fitted model in both $2\text{-}D$ (\texttt{model\_2d}), and $p\text{-}D$ (\texttt{model\_highd}), and triangular mesh (\texttt{trimesh\_data}).

@@

Constructing the $2\text{-}D$ model primarily involves (i) scaling the NLDR data, (ii) binning the data, (iii) obtaining bin centroids, (iv) connecting centroids with line segments to indicate neighbors, and (v) removing low-density hexagons.

@@

Finally, a tibble is returned containing a unique hexagon ID (\texttt{h}) along with the corresponding x and y centroid coordinates (\texttt{c\_x}, \texttt{c\_y}), which define the layout of the hexagonal grid over the $2\text{-}D$ space.

@@

After generating the centroids that define the hexagonal grid, the next step is to assign each point in the NLDR embedding to its nearest hexagonal bin. The \texttt{assign\_data()} function performs this assignment by calculating the $2\text{-}D$ Euclidean distance between each point in the $2\text{-}D$ embedding and all hexagon centroids.

First, the function extracts the first two dimensions of the scaled NLDR embedding, which represent the $2\text{-}D$ layout. It then selects the corresponding x and y coordinates of each hexagon's centroid.

@@

The result is a \texttt{tibble} of the scaled $2\text{-}D$ embedding with an additional \texttt{h} column, indicating the hexagonal bin to which each point belongs.

@@

The function first groups the input data by \texttt{h}, which represents the hexagon ID associated with each point in the $2\text{-}D$ layout. Within each group, it collects the \texttt{ID}s into a list, resulting in a summary where each row corresponds to a single hexagon. The resulting column, \texttt{pts\_list}, contains all point identifiers associated with that hexagon.

@@

To represent the neighborhood structure of hexagonal bins in a $2\text{-}D$ layout, we employ Delaunay triangulation \citep{lee1980, albrecht2024} on the centroids of hexagons. This geometric approach is used to infer which bins are considered neighbors.

@@

Not all hexagons contain meaningful information. Some may have very few or no data points due to the sparsity or shape of the underlying structure. Simply removing hexagons with low counts (e.g., fewer than a fixed threshold) can lead to gaps or ``holes'' in the $2\text{-}D$ structure, potentially disrupting the continuity of the representation.

@@

The final step involves lifting the fitted $2\text{-}D$ model into $p\text{-}D$. This is done by modelling a point in $p\text{-}D$ as the $p\text{-}D$ mean of data points in the $2\text{-}D$ centroid. This is performed using the \texttt{avg\_highd\_data()} function, which takes $p\text{-}D$ data (\texttt{highd\_data}) and embedding data with their corresponding hexagonal bin IDs as inputs (\texttt{scaled\_nldr\_hexid}).

@@

The \texttt{predict\_emb()} function is used to predict a point in a $2\text{-}D$ embedding for a new $p\text{-}D$ data point using the fitted model. This function is useful to predict $2\text{-}D$ embedding irrespective of the NLDR technique.

In the prediction process, first, the nearest $p\text{-}D$ model point is identified for the new $p\text{-}D$ data point by computing $p\text{-}D$ Euclidean distance. Then, the corresponding $2\text{-}D$ bin centroid mapping for the identified $p\text{-}D$ model point is determined. Finally, the coordinates of the identified $2\text{-}D$ bin centroid is used as the predicted NLDR embedding for the new $p\text{-}D$ data point.

@@

\subsection{Compute residuals and Root Within Bin Sum of Square (RWBSS)}\label{compute-residuals-and-root-within-bin-sum-of-square-rwbss}

Root Within Bin Sum of Square (RWBSS) are used as goodness of fit metrics for the model. These metrics can be computed using the \texttt{glance()} function, which provides a tidy output for evaluation.

The function requires both the fitted model object returned by \texttt{fit\_highd\_model()} and $p\text{-}D$ data to begin. The $p\text{-}D$ model output (\texttt{model\_highd}) is first renamed to avoid naming conflicts during subsequent data joins. It then uses the \texttt{predict\_emb()} function to assign each point in the $p\text{-}D$ dataset to a corresponding hexagon bin in the $2\text{-}D$ model, producing a prediction data frame that contains both the predicted bin assignment (\texttt{pred\_h}) and the original observation \texttt{ID}.

The function joins this prediction output with both the $p\text{-}D$ model and the $p\text{-}D$ data (to retrieve true coordinates). It then calculates squared differences between the original and predicted $p\text{-}D$ coordinates for each dimension, storing these as \texttt{error\_square\_x1}, \texttt{error\_square\_x2}, \ldots, up to the dimensionality of the data.

From these per-dimension errors, the function computes absolute error which is the sum of absolute differences across all dimensions and observations and the RWBSS which is the average of the total squared error per point.

These metrics are returned in a tibble as \texttt{Error} (absolute error) and \texttt{RWBSS} (root mean squared error). The computation of total absolute error and RWBSS is performed in C++ for efficiency using the internal \texttt{compute\_errors()} function.

@@

\begin{verbatim}
> # A tibble: 1 x 2
>   Error  RMSE
>   <dbl> <dbl>
> 1  196. 0.116
\end{verbatim}

Furthermore, \texttt{augment()} requires both the fitted model object returned by \texttt{fit\_highd\_model()} and $p\text{-}D$ data to begin. It extends the fitted model by adding prediction results and error diagnostics to the original $p\text{-}D$ data.

The function starts with the same process as is used in the \texttt{glance()} function to produce a predicted point in $p\text{-}D$ for each point in the $p\text{-}D$ dataset.

@@

The package offers several $2\text{-}D$ visualizations, including:

@@

The generated $p\text{-}D$ model, overlaid with the data, can also be visualized using \texttt{show\_langevitour}. Additionally, it features a function for visualizing the $2\text{-}D$ projection of the fitted model overlaid on the data, called \texttt{plot\_proj}.

@@

The \texttt{geom\_trimesh()} function introduces a custom \texttt{ggplot2} layer designed for visualizing $2\text{-}D$ wireframe on a provided set of bin centroids.

@@

\subsubsection{\texorpdfstring{$p\text{-}D$ model visualization}{model visualization}}\label{model-visualization}

To visualize how well the $p\text{-}D$ model captures the underlying structure of the high-dimensional data, we provide a tour of the model in $p\text{-}D$ using the \texttt{show\_langevitour()} function. This function renders a dynamic projection of both the high-dimensional data and the model using the \texttt{langevitour} R package \citep{paul2023}.

@@

As an alternative to \texttt{langevitour}, users can explore the fitted $p\text{-}D$ model using the \texttt{detourr} \citep{casper2025}. The combined data object from \texttt{comb\_data\_model()} can be passed directly to the \texttt{detour()} function, where \texttt{tour\_aes()} defines the projection variables and color mapping. The visualization is rendered using \texttt{show\_scatter()}, which can display both data points and the model's structural edges via the \texttt{edges} argument.

@@

In the resulting interactive visualization, blue points represent the high-dimensional data, orange points represent the model centroids from each bin, and the lines between model points reflect the $2\text{-}D$ wireframe structure mapped to high-dimensional space.

@@

To support interactive evaluation of how well the $p\text{-}D$ model captures the structure of the high-dimensional data, we introduce \texttt{show\_link\_plots()}. This visualization combines two complementary views: the nonlinear dimension reduction (NLDR) representation and a dynamic tour of the model ovelaid the data in the high-dimensional space. Both views are interactively linked, enabling users to explore.

Before visualization, the input data must be prepared using the \texttt{comb\_all\_data\_model()} function. This function combines the high-dimensional data (\texttt{highd\_data}), the NLDR data (\texttt{nldr\_data}), and the bin-averaged high-dimensional model representation (\texttt{model\_highd}) aligned to the $2\text{-}D$ bin layout (\texttt{model\_2d}):

@@

The function \texttt{show\_link\_plots()} generates two side-by-side, interactively linked plots; a $2\text{-}D$ NLDR representation, and a dynamic projection tour in the original high-dimensional space (using the \texttt{langevitour} package), displaying both the data and the model. The function takes the output from \texttt{comb\_all\_data\_model()} (\texttt{point\_data}) and \texttt{edge\_data} which defines connections between neighboring bins.

@@

Before visualization, the input data must be prepared using the \texttt{comb\_all\_data\_model\_error()} function. The function requires several arguments: points data which contain high-dimensional data (\texttt{highd\_data}), NLDR data (\texttt{nldr\_data}), high-dimensional model data (\texttt{model\_highd}), $2\text{-}D$ model data (\texttt{model\_2d}), and model error (\texttt{error\_data}).

@@

The function \texttt{show\_error\_link\_plots()} generates three side-by-side, interactively linked plots; a error distribution, a $2\text{-}D$ NLDR representation, and a dynamic projection tour in the original high-dimensional space (using the \texttt{langevitour} package), displaying both the data and the model. The function takes the output from \texttt{comb\_all\_data\_model\_error()} (\texttt{point\_data}) and \texttt{edge\_data} which defines connections between neighboring bins.

@@

As an alternative to using \texttt{langevitour}, link plots can also be generated with the \texttt{detourr} package. In this setup, users can manually construct the linked visualization using the \texttt{crosstalk} \citep{joe2023} and \texttt{htmltools} \citep{joe2024} packages. The interactive layout is created by arranging the $2\text{-}D$ NLDR plot (\texttt{nldr\_plt}), the optional error distribution plot (\texttt{error\_plt}), and the tour view produced with \texttt{detourr}, side by side within a flexible grid. The NLDR and error plots are rendered with \texttt{ggplotly()} \citep{chapman2020} to enable interactive linking. The \texttt{bscols()} function from \texttt{crosstalk} manages synchronization across these panels, allowing for linked brushing and coordinated selection between these interactive plots.

@@

Limb muscle cells of mice in \citet{tabula2018} are examined. There are \(1067\) single cells, with \(14997\) gene expressions. Following their pre-processing, different NLDR methods were performed using ten principal components. Figure \ref{fig:limb-rwbss} (b) is the reproduction of the published plot. The question is whether this accurately represents the cluster structure in the data. Our method help to provide a better $2\text{-}D$ layout.

@@

plot_rmse_layouts(plots = list(error_plot_limb, nldr1, 
                               nldr2, nldr3, nldr4, 
                               nldr5, nldr6), design = design)
\end{verbatim}

\begin{figure}[H]
\includegraphics[width=1\linewidth]{paper-quollr_files/figure-latex/limb-rwbss-1} \caption{Assessing which of the 6 NLDR layouts on the limb muscle data is the better representation using RWBSS for varying binwidth ($a_1$). Colour  used for the lines and points in the left plot and in the scatterplots represents NLDR layout (a-f). Layout d is perform well at large binwidth (where the binwidth is not enough to capture the data struture) and poorly as bin width decreases. Layout f is the best choice.\label{fig:limb-rwbss}}\label{fig:limb-rwbss}
\end{figure}

\begin{figure}[H]
\includegraphics[width=1\linewidth]{paper-quollr_files/figure-latex/model-limb-1} \caption{Compare the published $2\text{-}D$ layout (Figure \ref{fig:limb-rwbss}b) and the $2\text{-}D$ layout selected (Figure \ref{fig:limb-rwbss}f) by RWBSS plot (Figure \ref{fig:limb-rwbss}) from the tSNE, UMAP, PHATE, TriMAP, and PaCMAP with different hyper-parameters. The Limb muscle data ($n =  1067$) has seven close different shaped clusters in $10\text{-}D$.}\label{fig:model-limb}
\end{figure}

@@

The \texttt{quollr} package introduces a new framework for interpreting NLDR outputs by fitting a geometric wireframe model in $2\text{-}D$ and lifting it into high-dimensional space. This lifted model provides a direct way to assess how well a $2\text{-}D$ layout, produced by methods such as tSNE, UMAP, PHATE, TriMAP, or PaCMAP, preserves the structure of the original high-dimensional data. The approach offers both numerical and visual diagnostics to support the selection of NLDR methods and tuning hyper-parameters that produce the most accurate $2\text{-}D$ representations.

In contrast to the common practice of visually inspecting scatterplots for clusters or patterns, \texttt{quollr} provides a quantitative route for evaluation. It enables the computation of RWBSS and residuals between the original high-dimensional data and the lifted model, offering interpretable diagnostics. These diagnostics are complemented by interactive linked plots and high-dimensional dynamic visualizations using the \texttt{langevitour} package, allowing users to inspect where the model fits well and where it does not.

>>>>

This paper presents the R package, \texttt{quollr}, which is useful for understanding how NLDR warps high-dimensional space and fits the data. Starting with an NLDR layout, our approach is to create a \(2\text{-}D\) wireframe model representation, that can be lifted and displayed in the high-dimensional (\(p\text{-}D\)) space (Figure \ref{fig:overview}).

\begin{figure}[H]
\includegraphics[width=1\linewidth]{paper-quollr_files/figure-latex/overview-1} \caption{Wireframe model representation of the NLDR layout, lifted and displayed in high-dimensional space. The left panel shows the NLDR layout with a triangular mesh overlay, forming the wireframe structure. This mesh can be lifted into higher dimensions and projected to examine how the geometric structure of the data is preserved. Panels (a1–a4) display different $2 ext{-}D$ projections of the lifted wireframe, where the underlying curved sheet structure of the data is more clearly visible. The triangulated mesh highlights how local neighborhoods in the layout correspond to relationships in the high-dimensional space, enabling diagnostics of distortion and preservation across dimensions.}\label{fig:overview}
\end{figure}

@@

The package is organized into seven core modules corresponding to stages of the analysis workflow: preprocessing, \(2\text{-}D\) model construction, lifting into \(p\text{-}D\), prediction, error computation, visualization, and interactivity. Each module performs a distinct task and communicates through data objects.

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  Data preprocessing: The function \texttt{gen\_scaled\_data()} standardizes the embedding data, manage variable naming, and ensure consistent identifiers across high-dimensional and embedded datasets.
\item
  Construct \(2\text{-}D\) model: A series of functions \texttt{hex\_binning()}, \texttt{merge\_hexbin\_centroids()}, \texttt{tri\_bin\_centroids()}, \texttt{gen\_edges()}, and \texttt{update\_trimesh\_index()} generate the hexagonal grid, compute bin centroids, and connect the triangular mesh that defines local neighborhoods in the \(2\text{-}D\) space.
\item
  Lift the model into \(p\text{-}D\): The function \texttt{avg\_highd\_data()} computes the average of the high-dimensional variables for each bin, linking the \(2\text{-}D\) representation back to the original data space.
\item
  Prediction: The function \texttt{predict\_emb()} estimates the embedding of new high-dimensional observations based on the fitted model.
\item
  Error computation: The \texttt{glance()} and \texttt{augment()} function summarizes model performance by comparing the predicted and original embeddings.
\item
  Visualization: Functions such as \texttt{geom\_hexgrid()}, \texttt{geom\_trimesh()}, and \texttt{show\_langevitour()} provide tools for exploring the fitted models through static and dynamic visualizations.
\item
  Interactivity: The functions \texttt{comb\_all\_data\_model()} and \texttt{show\_link\_plots()} generate interactive linked visualizations that connect the \(2\text{-}D\) NLDR layout, the corresponding tour view, and the fitted model. Similarly, \texttt{comb\_all\_data\_model\_error()} and \texttt{show\_error\_link\_plots()} integrate the error distribution with the \(2\text{-}D\) embedding and dynamic high-dimensional view, enabling interactivity across multiple visual components.
\end{enumerate}

@@

\begin{itemize}
\item
  \texttt{scaled\_nldr\_obj}: the output of \texttt{gen\_scaled\_data()}, which rescales the embedding to the range \([0, 1] \times [0, y_{2,\text{max}}]\), where \(y_{2,\text{max}} = r_2 / r_1\) is the ratio of the embedding ranges. It includes the scaled coordinates (\texttt{scaled\_nldr}) and the original limits (\texttt{lim1}, \texttt{lim2}).
\item
  \texttt{hex\_bin\_obj}: the object created by \texttt{hex\_binning()}, which defines the structure of the two-dimensional hexagonal grid used in modeling. It includes the grid spacing (\texttt{a1}, \texttt{a2}), the number of bins along each axis, the centroids of all hexagons, polygon coordinates for plotting, and the mapping of each data point to its assigned hexagon.
\item
  \texttt{highd\_vis\_model}: the main model object returned by \texttt{fit\_highd\_model()}. It stores all components of the fitted visualization model, including the scaled NLDR data (\texttt{nldr\_scaled\_obj}), the hexagonal bin structure (\texttt{hb\_obj}), the averaged \(p\text{-}D\) summaries for each bin (\texttt{model\_highd}), the corresponding \(2\text{-}D\) bin centroids (\texttt{model\_2d}), and the triangulated mesh connecting neighboring bins (\texttt{trimesh\_data}).
\end{itemize}

@@

Several core computations within \texttt{quollr} are optimized using compiled C++ code via the \texttt{Rcpp} and \texttt{RcppArmadillo} packages. While the user interacts with high-level R functions, performance-critical steps such as nearest-neighbor searches (\texttt{compute\_highd\_dist()}), error metrics (\texttt{compute\_errors()}), \(2\text{-}D\) distance calculations (\texttt{calc\_2d\_dist\_cpp()}), and generation of hexagon coordinates (\texttt{gen\_hex\_coord\_cpp()}) are handled internally in C++. This design provides significant speedups when analyzing large datasets while maintaining a user-friendly R interface. These C++ functions are not exported but are bundled within the package and fully accessible for inspection in the source code.

@@

This function requires several parameters: the high-dimensional data (\texttt{highd\_data}), the emdedding data (\texttt{nldr\_data}), the number of bins along the x-axis (\texttt{b1}), the buffer amount as a proportion of data (\texttt{q}), and benchmark value to extract high density hexagons (\texttt{hd\_thresh}). The function returns an object of class \texttt{highd\_vis\_model} containing the scaled NLDR object (\texttt{nldr\_scaled\_obj}) with three elements: the first is the scaled NLDR data (\texttt{scaled\_nldr}), and the second and third are the limits of the original NLDR data (\texttt{lim1} and \texttt{lim2}); the hexagonal object (\texttt{hb\_obj}), the fitted model in both \(2\text{-}D\) (\texttt{model\_2d}), and \(p\text{-}D\) (\texttt{model\_highd}), and triangular mesh (\texttt{trimesh\_data}).

@@

Constructing the \(2\text{-}D\) model primarily involves (i) scaling the NLDR data, (ii) binning the data, (iii) obtaining bin centroids, (iv) connecting centroids with line segments to indicate neighbors, and (v) removing low-density hexagons.

@@

Finally, a tibble is returned containing a unique hexagon ID (\texttt{h}) along with the corresponding x and y centroid coordinates (\texttt{c\_x}, \texttt{c\_y}), which define the layout of the hexagonal grid over the \(2\text{-}D\) space.

@@

After generating the centroids that define the hexagonal grid, the next step is to assign each point in the NLDR embedding to its nearest hexagonal bin. The \texttt{assign\_data()} function performs this assignment by calculating the \(2\text{-}D\) Euclidean distance between each point in the \(2\text{-}D\) embedding and all hexagon centroids.

First, the function extracts the first two dimensions of the scaled NLDR embedding, which represent the \(2\text{-}D\) layout. It then selects the corresponding x and y coordinates of each hexagon's centroid.

@@

The result is a \texttt{tibble} of the scaled \(2\text{-}D\) embedding with an additional \texttt{h} column, indicating the hexagonal bin to which each point belongs.

@@

The function first groups the input data by \texttt{h}, which represents the hexagon ID associated with each point in the \(2\text{-}D\) layout. Within each group, it collects the \texttt{ID}s into a list, resulting in a summary where each row corresponds to a single hexagon. The resulting column, \texttt{pts\_list}, contains all point identifiers associated with that hexagon.

@@

To represent the neighborhood structure of hexagonal bins in a \(2\text{-}D\) layout, we employ Delaunay triangulation \citep{lee1980, albrecht2024} on the centroids of hexagons. This geometric approach is used to infer which bins are considered neighbors.

@@

Not all hexagons contain meaningful information. Some may have very few or no data points due to the sparsity or shape of the underlying structure. Simply removing hexagons with low counts (e.g., fewer than a fixed threshold) can lead to gaps or ``holes'' in the \(2\text{-}D\) structure, potentially disrupting the continuity of the representation.

@@

The final step involves lifting the fitted \(2\text{-}D\) model into \(p\text{-}D\). This is done by modelling a point in \(p\text{-}D\) as the \(p\text{-}D\) mean of data points in the \(2\text{-}D\) centroid. This is performed using the \texttt{avg\_highd\_data()} function, which takes \(p\text{-}D\) data (\texttt{highd\_data}) and embedding data with their corresponding hexagonal bin IDs as inputs (\texttt{scaled\_nldr\_hexid}).

@@

The \texttt{predict\_emb()} function is used to predict a point in a \(2\text{-}D\) embedding for a new \(p\text{-}D\) data point using the fitted model. This function is useful to predict \(2\text{-}D\) embedding irrespective of the NLDR technique.

In the prediction process, first, the nearest \(p\text{-}D\) model point is identified for the new \(p\text{-}D\) data point by computing \(p\text{-}D\) Euclidean distance. Then, the corresponding \(2\text{-}D\) bin centroid mapping for the identified \(p\text{-}D\) model point is determined. Finally, the coordinates of the identified \(2\text{-}D\) bin centroid is used as the predicted NLDR embedding for the new \(p\text{-}D\) data point.

@@

\subsection{Compute residuals and hexbin error (HBE)}\label{compute-residuals-and-hexbin-error-hbe}

Hexbin error (HBE) are used as goodness of fit metrics for the model. These metrics can be computed using the \texttt{glance()} function, which provides a tidy output for evaluation.

The function requires both the fitted model object returned by \texttt{fit\_highd\_model()} and \(p\text{-}D\) data to begin. The \(p\text{-}D\) model output (\texttt{model\_highd}) is first renamed to avoid naming conflicts during subsequent data joins. It then uses the \texttt{predict\_emb()} function to assign each point in the \(p\text{-}D\) dataset to a corresponding hexagon bin in the \(2\text{-}D\) model, producing a prediction data frame that contains both the predicted bin assignment (\texttt{pred\_h}) and the original observation \texttt{ID}.

The function joins this prediction output with both the \(p\text{-}D\) model and the \(p\text{-}D\) data (to retrieve true coordinates). It then calculates squared differences between the original and predicted \(p\text{-}D\) coordinates for each dimension, storing these as \texttt{error\_square\_x1}, \texttt{error\_square\_x2}, \ldots, up to the dimensionality of the data.

From these per-dimension errors, the function computes absolute error which is the sum of absolute differences across all dimensions and observations and the HBE which is the average of the total squared error per point.

These metrics are returned in a tibble as \texttt{Error} (absolute error) and \texttt{HBE} (root mean squared error). The computation of total absolute error and HBE is performed in C++ for efficiency using the internal \texttt{compute\_errors()} function.

@@

\begin{verbatim}
> # A tibble: 1 x 2
>   Error   HBE
>   <dbl> <dbl>
> 1  196. 0.116
\end{verbatim}

Furthermore, \texttt{augment()} requires both the fitted model object returned by \texttt{fit\_highd\_model()} and \(p\text{-}D\) data to begin. It extends the fitted model by adding prediction results and error diagnostics to the original \(p\text{-}D\) data.

The function starts with the same process as is used in the \texttt{glance()} function to produce a predicted point in \(p\text{-}D\) for each point in the \(p\text{-}D\) dataset.

@@

The package offers several \(2\text{-}D\) visualizations, including:

@@

The generated \(p\text{-}D\) model, overlaid with the data, can also be visualized using \texttt{show\_langevitour}. Additionally, it features a function for visualizing the \(2\text{-}D\) projection of the fitted model overlaid on the data, called \texttt{plot\_proj}.

@@

The \texttt{geom\_trimesh()} function introduces a custom \texttt{ggplot2} layer designed for visualizing \(2\text{-}D\) wireframe on a provided set of bin centroids.

@@

\subsubsection{\texorpdfstring{\(p\text{-}D\) model visualization}{p\textbackslash text\{-\}D model visualization}}\label{ptext-d-model-visualization}

To visualize how well the \(p\text{-}D\) model captures the underlying structure of the high-dimensional data, we provide a tour of the model in \(p\text{-}D\) using the \texttt{show\_langevitour()} function. This function renders a dynamic projection of both the high-dimensional data and the model using the \texttt{langevitour} R package \citep{paul2023}.

@@

As an alternative to \texttt{langevitour}, users can explore the fitted \(p\text{-}D\) model using the \texttt{detourr} \citep{casper2025}. The combined data object from \texttt{comb\_data\_model()} can be passed directly to the \texttt{detour()} function, where \texttt{tour\_aes()} defines the projection variables and color mapping. The visualization is rendered using \texttt{show\_scatter()}, which can display both data points and the model's structural edges via the \texttt{edges} argument.

@@

In the resulting interactive visualization, blue points represent the high-dimensional data, orange points represent the model centroids from each bin, and the lines between model points reflect the \(2\text{-}D\) wireframe structure mapped to high-dimensional space.

@@

To support interactive evaluation of how well the \(p\text{-}D\) model captures the structure of the high-dimensional data, we introduce \texttt{show\_link\_plots()}. This visualization combines two complementary views: the nonlinear dimension reduction (NLDR) representation and a dynamic tour of the model ovelaid the data in the high-dimensional space. Both views are interactively linked, enabling users to explore.

Before visualization, the input data must be prepared using the \texttt{comb\_all\_data\_model()} function. This function combines the high-dimensional data (\texttt{highd\_data}), the NLDR data (\texttt{nldr\_data}), and the bin-averaged high-dimensional model representation (\texttt{model\_highd}) aligned to the \(2\text{-}D\) bin layout (\texttt{model\_2d}):

@@

The function \texttt{show\_link\_plots()} generates two side-by-side, interactively linked plots; a \(2\text{-}D\) NLDR representation, and a dynamic projection tour in the original high-dimensional space (using the \texttt{langevitour} package), displaying both the data and the model. The function takes the output from \texttt{comb\_all\_data\_model()} (\texttt{point\_data}) and \texttt{edge\_data} which defines connections between neighboring bins.

@@

Before visualization, the input data must be prepared using the \texttt{comb\_all\_data\_model\_error()} function. The function requires several arguments: points data which contain high-dimensional data (\texttt{highd\_data}), NLDR data (\texttt{nldr\_data}), high-dimensional model data (\texttt{model\_highd}), \(2\text{-}D\) model data (\texttt{model\_2d}), and model error (\texttt{error\_data}).

@@

The function \texttt{show\_error\_link\_plots()} generates three side-by-side, interactively linked plots; a error distribution, a \(2\text{-}D\) NLDR representation, and a dynamic projection tour in the original high-dimensional space (using the \texttt{langevitour} package), displaying both the data and the model. The function takes the output from \texttt{comb\_all\_data\_model\_error()} (\texttt{point\_data}) and \texttt{edge\_data} which defines connections between neighboring bins.

@@

As an alternative to using \texttt{langevitour}, link plots can also be generated with the \texttt{detourr} package. In this setup, users can manually construct the linked visualization using the \texttt{crosstalk} \citep{joe2023} and \texttt{htmltools} \citep{joe2024} packages. The interactive layout is created by arranging the \(2\text{-}D\) NLDR plot (\texttt{nldr\_plt}), the optional error distribution plot (\texttt{error\_plt}), and the tour view produced with \texttt{detourr}, side by side within a flexible grid. The NLDR and error plots are rendered with \texttt{ggplotly()} \citep{chapman2020} to enable interactive linking. The \texttt{bscols()} function from \texttt{crosstalk} manages synchronization across these panels, allowing for linked brushing and coordinated selection between these interactive plots.

@@

Limb muscle cells of mice in \citet{tabula2018} are examined. There are \(1067\) single cells, with \(14997\) gene expressions. Following their pre-processing, different NLDR methods were performed using ten principal components. Figure \ref{fig:limb-hbe} (b) is the reproduction of the published plot. The question is whether this accurately represents the cluster structure in the data. Our method help to provide a better \(2\text{-}D\) layout.

@@

plot_hbe_layouts(plots = list(error_plot_limb, nldr1, 
                             nldr2, nldr3, nldr4, 
                             nldr5, nldr6), design = design)
\end{verbatim}

\begin{figure}[H]
\includegraphics[width=1\linewidth]{paper-quollr_files/figure-latex/limb-hbe-1} \caption{Assessing which of the 6 NLDR layouts on the limb muscle data is the better representation using HBE for varying binwidth ($a_1$). Colour  used for the lines and points in the left plot and in the scatterplots represents NLDR layout (a-f). Layout d is perform well at large binwidth (where the binwidth is not enough to capture the data struture) and poorly as bin width decreases. Layout f is the best choice.\label{fig:limb-hbe}}\label{fig:limb-hbe}
\end{figure}

\begin{figure}[H]
\includegraphics[width=1\linewidth]{paper-quollr_files/figure-latex/model-limb-1} \caption{Compare the published $2\text{-}D$ layout (Figure \ref{fig:limb-hbe}b) and the $2\text{-}D$ layout selected (Figure \ref{fig:limb-hbe}f) by HBE plot (Figure \ref{fig:limb-hbe}) from the tSNE, UMAP, PHATE, TriMAP, and PaCMAP with different hyper-parameters. The Limb muscle data ($n =  1067$) has seven close different shaped clusters in $10\text{-}D$.}\label{fig:model-limb}
\end{figure}

@@

The \texttt{quollr} package introduces a new framework for interpreting NLDR outputs by fitting a geometric wireframe model in \(2\text{-}D\) and lifting it into high-dimensional space. This lifted model provides a direct way to assess how well a \(2\text{-}D\) layout, produced by methods such as tSNE, UMAP, PHATE, TriMAP, or PaCMAP, preserves the structure of the original high-dimensional data. The approach offers both numerical and visual diagnostics to support the selection of NLDR methods and tuning hyper-parameters that produce the most accurate \(2\text{-}D\) representations.

In contrast to the common practice of visually inspecting scatterplots for clusters or patterns, \texttt{quollr} provides a quantitative route for evaluation. It enables the computation of HBE and residuals between the original high-dimensional data and the lifted model, offering interpretable diagnostics. These diagnostics are complemented by interactive linked plots and high-dimensional dynamic visualizations using the \texttt{langevitour} package, allowing users to inspect where the model fits well and where it does not.
